Indicators for Ecosystem Condition in Norway

### 2024-09-27 gjengroingMetaModel.Rmd
Test

Zander Venter

September 2023

@@

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(tidyverse)
library(gridExtra)
library(RColorBrewer)
library(flextable)
# Set global variable defining whether you want to run everything from scratch (very long runtime) or use pre-exported data (short runtime)
runFromScratch <- FALSE

@@

```{r, echo=F}
Ecosystem <- "Våtmark og Semi-/Naturlig-åpne" 
Egenskap  <- "Funksjonell sammensetning innen trofiske nivåer"
ECT       <- "Structural state characteristic" 
Contact   <- "Zander Venter"
```

@@

The Norwegian word "gjengroing" is directly translated to "regrowing" in English. The gjengroing indicator describes the regrowth of woody vegetation (trees and bushes) in open ecosystems (wetland, semi- and naturally open areas) across Norway. We will use a spatial reference approach where reference areas define good or optimal vegetation regrowth heights.

The workflow spans two platforms including RStudio and Google Earth Engine (GEE). To reproduce this workflow you will need a GEE account and access to the NINA RStudio and R/GeoSpatialData servers.

@@

#### Regions {#reg-enc}

>>>>

Andrew Gray

September 2024

@@

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(tidyverse)
library(gridExtra)
library(RColorBrewer)
library(flextable)
library(terra)
library(dplyr)
library(parallel)
library(stringr)
library(httr)
library(jsonlite)
# Set global variable defining whether you want to run everything from scratch (very long runtime) or use pre-exported data (short runtime)
runFromScratch <- FALSE

@@

```{r, echo=F}
Ecosystem <- "Våtmark og Semi-/Naturlig-åpne" 
Egenskap  <- "Funksjonell sammensetning innen trofiske nivåer"
ECT       <- "Structural state characteristic" 
Contact   <- "Andrew Gray"
```

@@

The Norwegian word "gjengroing" is directly translated to "regrowing" in English. The gjengroing indicator describes the regrowth of woody vegetation (trees and bushes) in open ecosystems (wetland, semi- and naturally open areas) across Norway. Here, we use meta's canopy height base model to predict canopy height within areas of interest using airborne imagery from norge i bilder. This approach will allow future monitoring of encroachment, as airborne imagery is scheduled to be collected every 5 - 10 years whereas national LiDAR data has no future schedule. It is worth noting that the yet-to-be-launched P-Band radar satellite, Biomass, may provide an more precise method for future monitoring once data is available. Here though, we will use a spatial reference approach where reference areas define good or optimal vegetation regrowth heights.

The workflow spans two platforms including RStudio and Python.

@@

##### Directories

```{r}
wms_base_url <- 'https://wms.geonorge.no/skwms1/wms.nib?SERVICE=WMS&VERSION=1.1.1&REQUEST=GetMap&SRS=EPSG:25833'
output_dir <- "/data/P-Prosjekter2/412421_okologisk_tilstand_2024/Andrew/meta/HighResCanopyHeight-main/DataDownload/testHarstad/tiles"
metadata_url <- 'https://tjenester.norgeibilder.no/rest/projectMetadata.ashx'
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```

###### Querying the airborne imagery metadata

This enables us to determine the date of the orthophoto tiles we will later download from the geonorge webserver.

```{r}
 query_metadata <- function(xmin, ymin, xmax, ymax, crs_code) {
  bbox_coords <- sprintf("%f,%f;%f,%f;%f,%f;%f,%f", xmin, ymin, xmin, ymax, xmax, ymax, xmax, ymin)

  param <- list(
    Filter = "ortofototype in (1,2,3,4,5,7,8,9,10,11,12)",
    Coordinates = bbox_coords,
    InputWKID = as.character(crs_code),
    ReturnMetadata = TRUE,
    StopOnCover = FALSE
  )

  response <- httr::GET(
    url = metadata_url,
    query = list(request = jsonlite::toJSON(param, auto_unbox = TRUE))
  )

  response_content <- httr::content(response, "text", encoding = "UTF-8")

  if (httr::http_status(response)$category != "Success") {
    return(NULL)
  }

  if (response_content == "" || is.null(response_content)) {
    return(NULL)
  }

  parsed_metadata <- tryCatch({
    fromJSON(response_content, flatten = TRUE)
  }, error = function(e) {
    return(NULL)
  })

  return(parsed_metadata)
}
```
The WMS provides the most recent image over each search area so we can pull the date from the metadata service to add to the tile filename.

```{r}
get_most_recent_metadata <- function(metadata) {
  filtered_metadata <- metadata$ProjectMetadata %>%
    filter(!is.na(properties.fotodato_date)) %>%
    mutate(fotodato_date = as.Date(properties.fotodato_date)) %>%
    arrange(desc(fotodato_date)) %>%
    slice(1)  # Get the most recent one

  if (nrow(filtered_metadata) == 0) {
    return(NULL)
  }

  return(filtered_metadata)
}
```

##### Defining the area of interest
Here we include a shapefile of our area of intertest. The script will loop through all features in the shapefile, calculate their extent and query the WMS to download orthophotos within that extent in 256 x 256 pixel tiles. 
```{r}
ROIs <- vect("/data/P-Prosjekter2/412421_okologisk_tilstand_2024/Andrew/meta/HighResCanopyHeight-main/DataDownload/testHarstad/aoi.shp")
RegionPolygonUTM33N <- st_as_sf(ROIs)
if (st_crs(RegionPolygonUTM33N)$epsg != 25833) {
  RegionPolygonUTM33N <- st_transform(RegionPolygonUTM33N, crs = 25833)
}

tile_size <- 256
resolution <- 0.5  # 0.5 meter resolution
```

This is the function to get and download the orthophotos over our areas of interest. 
```{r}
fn.getOrtoImages <- function(geometry, id_field, output_dir) {
  tryCatch({
    if (!st_is_valid(geometry)) {
      geometry <- st_make_valid(geometry)
    }

    geom_extent <- st_bbox(geometry)

    # Query metadata to get the most recent layer and date
    metadata <- query_metadata(geom_extent$xmin, geom_extent$ymin, geom_extent$xmax, geom_extent$ymax, 25833)

    if (is.null(metadata) || is.null(metadata$ProjectMetadata)) {
      return(NULL)
    }

    recent_project <- get_most_recent_metadata(metadata)

    if (is.null(recent_project)) {
      return(NULL)
    }

    recent_layer <- recent_project$properties.prosjektnavn
    image_date <- recent_project$properties.fotodato_date

    tile_width_meters <- tile_size * resolution
    tile_height_meters <- tile_size * resolution

    # Ensure tile size remains 256x256
    x_min <- floor(geom_extent$xmin / tile_width_meters) * tile_width_meters
    y_min <- floor(geom_extent$ymin / tile_height_meters) * tile_height_meters
    x_max <- ceiling(geom_extent$xmax / tile_width_meters) * tile_width_meters
    y_max <- ceiling(geom_extent$ymax / tile_height_meters) * tile_height_meters

    # Calculate number of tiles in x and y directions
    x_tiles <- ceiling((x_max - x_min) / tile_width_meters)
    y_tiles <- ceiling((y_max - y_min) / tile_height_meters)

    tile_index <- 1

    for (i in 0:(x_tiles - 1)) {
      for (j in 0:(y_tiles - 1)) {
        xmin <- x_min + i * tile_width_meters
        xmax <- xmin + tile_width_meters
        ymin <- y_min + j * tile_height_meters
        ymax <- ymin + tile_height_meters

        # Ensure the tiles have consistent size
        bbox_polygon <- st_polygon(list(matrix(c(
          xmin, ymin,
          xmin, ymax,
          xmax, ymax,
          xmax, ymin,
          xmin, ymin
        ), ncol = 2, byrow = TRUE)))
        bbox_polygon <- st_sfc(bbox_polygon, crs = st_crs(geometry))

        bbx <- st_bbox(bbox_polygon)

        # Construct WMS URL dynamically with correct SRS, LAYERS, and BBOX in EPSG:25833
        wms_url <- paste0(
          'https://wms.geonorge.no/skwms1/wms.nib?',
          'SERVICE=WMS&VERSION=1.1.1&REQUEST=GetMap',
          '&LAYERS=', URLencode(recent_layer),
          '&SRS=EPSG:25833',
          '&BBOX=', paste(bbx[c("xmin", "ymin", "xmax", "ymax")], collapse = ","),
          '&WIDTH=', tile_size,
          '&HEIGHT=', tile_size,
          '&FORMAT=image/tiff',
          '&TRANSPARENT=TRUE'
        )

        # Output file path using the "id" field, tile index, and image date in the file name
        t <- file.path(output_dir, paste0(image_date,"_", id_field, "_", tile_index, ".tif"))

        # GDAL translate options for 0.5 meter pixel size
        reso <- c('-tr', as.character(resolution), as.character(resolution), '-co', 'COMPRESS=NONE')

        # Try downloading the tile
        result <- try({
          sf::gdal_utils('translate', source = wms_url, destination = t, options = reso)
        }, silent = TRUE)

        # Retry mechanism
        for (k in 1:5) {
          if (!file.exists(t)) {
            Sys.sleep(1)
            result <- try({
              sf::gdal_utils('translate', source = wms_url, destination = t, options = reso)
            }, silent = TRUE)
          }
        }

        tile_index <- tile_index + 1
      }
    }

  }, error = function(e) {
    return(NULL)
  })
}
```

We can run the functions in parallel to speed up processing time adjust the number of cores to suit your setup.
```{r}
numCores <- 30

# Use mclapply for parallel processing
mclapply(1:nrow(RegionPolygonUTM33N), function(i) {
  geometry <- RegionPolygonUTM33N[i, ]
  id_field <- RegionPolygonUTM33N$id[i]  # Adjust field name if necessary
  fn.getOrtoImages(geometry, id_field, output_dir)
}, mc.cores = numCores)
```

######Zanders/Ida's script######
#### Regions {#reg-enc}
